\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-29 papers/jfp/7-experience.tex
More work tidying up S8
Drop some TODOs
Polish S7-S9

As a set of authors, we've written a paper about the \Shake build system \cite{mitchell2012shake},
applying \Shake to building GHC \cite{hadrian}, and a paper about the categorisation of build
systems \cite{mokhov2018buildsystems}. The original motivation of the research in that final paper was to create
``\Cloud \Shake'', and apply it to the GHC build system. In this section we tie these papers together,
reflecting on our experience since these papers.

@@

The original design of \Shake hasn't changed since the initial paper, but the implementation has continued to mature -- there have been roughly 5,000 subsequent commits to the \Shake project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add concepts like resources (if two rules contend on a single external resource), rewriting serialisation to be faster, documentation including a website\footnote{\url{https://shakebuild.com}}, and add lots of tests. The biggest change in that time period was moving from blocking threads to continuations for the suspending scheduler. The most visible change is switching from \hs{*>} to \hs{%>} for defining rules\footnote{A conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost all external and internal details remain the same.

We consider the lack of change suggestive that \Shake is based on fundamental principles -- principles we can now name and describe as a consequence of \cite{mokhov2018buildsystems}. There are two main aspects to the original \Shake paper \cite{mitchell2012shake} that are described more clearly in this paper, Firstly, the rebuilder can now be described as verifying step traces as per \S\ref{sec-step-traces}, with a much clearer relationship to the unoptimised verifying traces of \S\ref{sec-verifying-traces}. The second is the power of \Shake tasks (called actions in the original paper). Originally these tasks were described using the data type:

\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v r = Finished r
                  | Depends v (v -> Action k v r)
\end{minted}

This type is equivalent to \hs{Task Monad} -- the two types form an isomorphism. The original paper uses concrete types \hs{Key} and \hs{Value}, we generalise these types to \hs{k} and \hs{v}, and also add \hs{r} so that \hs{Action k v} can be an instance of \hs{Monad}. We define that \hs{Monad} instance, and use it to write conversion functions to/from our \hs{Task} type, witnessing the isomorhpism:

\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> f =<< op v

@@

fromAction :: Action k v v -> Task Monad k v
fromAction x = Task $ \fetch -> f fetch x
    where f fetch (Finished v) = return v
          f fetch (Depends d op) = f fetch . op =<< fetch d
\end{minted}

@@

\begin{minted}[xleftmargin=10pt]{haskell}
data Rule k v r = Rule {depends :: [k], action :: [v] -> r}
\end{minted}

Once again, this type is isomorphic to \hs{Task Applicative}, and we can define a similar \hs{Applicative} instance and conversion functions. By describing these types using \hs{Task} we are able to describe the differences more concisely (\hs{Monad} vs \hs{Applicative}) and use existing literature to determine what is and isn't possible.

@@

\todo{AM: Can you flesh this out}

@@

One of the benefits of using Shake is that we have access to high quality profiling information, allowing us to compute critical paths and other metrics (see \cite{mitchell2019ghcrebuildtimes} for a tour of the metrics). This information has shown us that GHC is slow to build (40m on an 8 CPU machine on Windows), that more CPUs would not help (on unlimited CPUs it would take at least 37m), and that a handful of steps (two Haskell compiles, some calls to \hs{configure}) are responsible for a significant amount of that time (at least 10m).

@@

Converting \Shake into \Cloud \Shake wasn't a difficult process once armed with a roadmap from \cite{mokhov2018buildsystems}. The key was the introduction of two new functions:

\begin{minted}[xleftmargin=10pt]{haskell}
addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]]
         -> v -> [FilePath] -> IO ()
lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver
            -> m (Maybe (v, [[k]], IO ()))
\end{minted}

These functions are suspiciously like \hs{recordCT} and \hs{constructCT} from \S\ref{sec-constructive-traces}, with their differences perhaps the most illustrative of the changes required. (We have made some minor changes from actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant differences.)

\begin{itemize}
\item There are two \hs{Ver} arguments being passed to each function. These are the versions of the build script, and the rule for this particular key. If either version changes then it is as though the key has changed, and nothing will match. These versions are important to avoid using stale build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than a simple list. The reason is that \Shake allows a list of dependencies to be supplied at once, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the files that this rule produces. These produced files include those which are output keys from a rule and those declared with \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a list. This change was made for simplicity.
\end{itemize}

To integrate these functions into \Shake we found the most expedient route was to leave \Shake with verifying traces, but if the verifying trace doesn't match, we consult the constructive trace. By bolting constructive traces onto the side of Shake we avoid reengineering of the central database. We haven't found any significant downsides from the bolt-on approach thus far, so it may be a sensible route to go even if developing from scratch -- allowing an optimised verified trace implementation in many cases, and falling back to a more complex implementation (requiring consulting remote servers) only rarely.

The one thing we haven't yet completed on the engineering side is a move to hosting caches over HTTP. At the moment all caches are on shared file systems. This approach can use mounted drives to mirror HTTP connections onto file systems, and reuse tools for managing file systems, share caches with rsync, and is simple. Unfortunately, on certain OS's (e.g. Windows) mounting an HTTP endpoint as a file system requires administrator privileges, so an HTTP cache is still desirable.

@@

\begin{quote}
Thanks for the symlinks release, we just finished upgrading this build system to use --share. ...
Building from scratch with a warm cache takes around 5 seconds, saving us up to 2 hours. Not bad!
\end{quote}

We have also got the GHC build working with \Cloud \Shake, resulting in build times of N minutes instead of M minutes. Due to aspects like \hs{configure} we don't see the dramatic reductions reported by Standard Chartered. Converting to a build suitable for sharing is not overly onerous, but nor is it trivial. Some things that require attention:

\begin{description}
\item[Irrelevant differences] A common problem is that you don't get shared caching when you want it. To get caching, it is important that all machines are identical for the aspects that involve the build system. In particular, that means that all paths must be relative, and all dependencies on system binaries (e.g. \hs{gcc}) should either be controlled by the build system (e.g. using \Nix) or have tolerence to account for minor machine variations.
\item[Insufficient produced files] A build rule must declare which files it produces, so these are stored in the cache and can be restored from the cache. If you don't use a remote cache then it is permissible to have files not known to the build system which change at the same time as files that are known. Once using a remote cache any file that is produced as a side-effect of a rule and then later used \emph{must} be declared. Failure to declare such files means that if some step produces a file, and a later step relies on it, that step breaks. In practice, most issues encountered during the move to Cloud builds for GHC were a lack of declared produced files.
\item[Insufficient dependencies] While insufficient dependencies are often a problem, the move to a Cloud build makes them even more critical, as missing dependencies are likely to mean a result is \emph{never built locally}, rather than just build locally only once. Furthermore, dependencies in a Cloud setting must be more complete. For example, using the build system from Figure \ref{fig-make}, with a local build system without early cutoff, if the rule to produce \hs{main.exe} also made use of \hs{main.c} but didn't declare so, that isn't a problem. With the addition of early cutoff, if \hs{main.c} changes in an important way but \hs{main.o} doesn't update that causes issues, but is a somewhat rare occurance. With Cloud builds that becomes somewhat more common.
\end{description}

To help with the final two issues -- insufficient dependencies and produced files -- we have further enhanced the \Shake lint modes, coupling them to a program called \hs{FSATrace}, which detects which files are read/written by a command line execution. Such information has been very helpful in making the GHC build Cloud ready.

>>>>

As a set of authors, we've written a paper about the \Shake build
system~\cite{mitchell2012shake}, applying \Shake to building GHC~\cite{hadrian},
and a paper about the categorisation of build
systems~\cite{mokhov2018buildsystems}. The original motivation of the research
in that final paper was to create ``\Cloud \Shake'', and apply it to the GHC
build system. In this section we tie these papers together, reflecting on our
experience since these papers.

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add
concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{\url{https://shakebuild.com}}, and
add a lot of tests. The biggest change in that time period was moving from
blocking threads to continuations for the suspending scheduler. The most visible
change is switching from \hs{*>} to \hs{@\%@>} for defining rules\footnote{A
conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost
all external and internal details remain the same.

We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of the categorisation paper~\cite{mokhov2018buildsystems}. There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. The second
aspect is the power of \Shake tasks, or ``actions'' in the original paper,
which were originally described using this data type:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{1mm}

\noindent
This type is equivalent to \hs{Task}~\hs{Monad} -- the two types form an
isomorphism. The original paper uses concrete types \hs{Key} and \hs{Value}, we
generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}. We define that
\hs{Monad} instance, and use it to write conversion functions to/from our
\hs{Task} type, witnessing the isomorphism:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f

@@

fromAction :: Action k v v -> Task Monad k v
fromAction x = Task $ \fetch -> f fetch x
  where
    f _     (Finished v  ) = return v
    f fetch (Depends d op) = fetch d >>= f fetch . op
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Rule k v a = Rule { depends :: [k], action :: [v] -> a }
\end{minted}
\vspace{1mm}

Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, this
type is isomorphic to \hs{Task}~\hs{Applicative}, and we can define a similar
\hs{Applicative} instance and conversion functions. By describing these types
using \hs{Task} we are able to describe the differences more concisely
(\hs{Monad} vs \hs{Applicative}), use existing literature to determine what is
and isn't possible, and explore other constraints beyond just \hs{Monad} and
\hs{Applicative}.

@@

\todo{AM}{Finish this section.}

@@

One of the benefits of using Shake is that we have access to high quality
profiling information, allowing us to compute critical paths and other metrics,
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for a tour of the metrics.
This information has shown us that GHC is slow to build (40m on an 8 CPU machine
on Windows), that more CPUs would not help (on unlimited CPUs it would take at
least 37m), and that a handful of steps (two Haskell compiles, some calls to
\cmd{configure}) are responsible for a significant amount of that time (at least
10m).

@@

Converting \Shake into \Cloud \Shake was not a difficult process once armed with
a roadmap~\cite{mokhov2018buildsystems}. The key was the introduction of two new
functions:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]]
         -> v -> [FilePath] -> IO ()
lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver
            -> m (Maybe (v, [[k]], IO ()))
\end{minted}
\vspace{1mm}

\noindent
These functions are suspiciously like \hs{recordCT} and \hs{constructCT}
from~\S\ref{sec-constructive-traces}, with their differences perhaps the most
illustrative of the changes required. (We have made some minor changes from
actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant
differences.)

\begin{itemize}
\item There are two \hs{Ver} arguments being passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be supplied at once, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

To integrate these functions into \Shake we found the most expedient route was
to leave \Shake with verifying traces, but if the verifying trace does not
match, we consult the constructive trace. By bolting constructive traces onto
the side of Shake we avoid re-engineering of the central database. We have not
found any significant downsides from the bolt-on approach thus far, so it may be
a sensible route to go even if developing from scratch -- allowing an optimised
verified trace implementation in many cases, and falling back to a more complex
implementation (requiring consulting remote servers) only rarely.

The one thing we have not yet completed on the engineering side is a move to
hosting caches over HTTP. At the moment all caches are on shared file systems.
This approach can use mounted drives to mirror HTTP connections onto file
systems, and reuse tools for managing file systems, share caches with
\cmd{rsync}, and is simple. Unfortunately, on certain operating systems (e.g.
Windows) mounting an HTTP endpoint as a file system requires administrator
privileges, so an HTTP cache is still desirable.

@@

\vspace{1mm}
\begin{center}
\parbox{0.8 \textwidth}{\emph{Thanks for the symlinks release, we just finished upgrading this build
system to use \cmd{-}\cmd{-share}. ... Building from scratch with a warm cache
takes around 5 seconds, saving us up to 2 hours. Not bad!}}
\end{center}
\vspace{1mm}

We have also got the GHC build working with \Cloud \Shake, resulting in build
times of 3m and 30s instead of 1h and 8m (on Windows). Due to aspects like slow
and fully sequential \cmd{configure} we do not see the dramatic reductions
reported by Standard Chartered. Converting to a build suitable for sharing is
not overly onerous, but nor is it trivial. Some things that require attention:

\begin{itemize}
\item \textbf{Irrelevant differences}: A common problem is that you do not get
      shared caching when you want it. To get caching, it is important that all
      machines are identical for the aspects that involve the build system. In
      particular, that means that all paths must be relative, and all
      dependencies on system binaries (e.g. \cmd{gcc}) should either be
      controlled by the build system (e.g. using \Nix) or have tolerance to
      account for minor machine variations.
\item \textbf{Insufficient produced files}: A build rule must declare all files
      it produces, so these are stored in the cache and can be restored from the
      cache. If you do not use a remote cache then it is permissible to have
      files not known to the build system which change at the same time as files
      that are known. Once using a remote cache any file that is produced as a
      side-effect of a rule and then later used \emph{must} be declared. Failure
      to declare such files means that if some step produces a file, and a later
      step relies on it, that step breaks. In practice, most issues encountered
      during the move to cloud builds for GHC were a lack of declared produced
      files.
\item \textbf{Insufficient dependencies}: While insufficient dependencies are
      often a problem, the move to a cloud build makes them even more critical,
      as missing dependencies are likely to mean a result is \emph{never built
      locally}, rather than just build locally only once. Furthermore,
      dependencies in a cloud setting must be more complete. For example, using
      the build system from Fig.~\ref{fig-make}, with a local build system
      without early cutoff, if the rule to produce \cmd{main.exe} also made use
      of \cmd{main.c} but did not declare so, that is not a problem. With the
      addition of early cutoff, if \cmd{main.c} changes in an important way but
      \cmd{main.o} is not rebuilt that causes issues, but is a somewhat rare
      occurrence. With cloud builds that becomes somewhat more common.
\end{itemize}

To help with the final two issues -- insufficient dependencies and produced
files -- we have further enhanced the \Shake lint modes, coupling them to a
program called \cmd{FSATrace}, which detects which files are read/written by a
command line execution. Such information has been very helpful in making the GHC
build cloud ready.

### 2019-04-29 papers/jfp/8-engineering.tex
Drop some TODOs
Polish S7-S9

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, if a the calcuation order is persisted over
successive runs (as \Excel does), potentially racey dependencies will be separated,
giving better parallelism over time.

\item The \hs{suspending} scheduler can be made parallel by starting multiple
dependencies simulataneously. One approach is to make the request for dependencies
take a list of keys, as implemented by \Shake. Another approach is to treat the
\hs{Applicative} dependencies of a \hs{Task Monad} in parallel, using the strategy
described by~\cite{marlow2014haxl}.
\end{enumerate}

Once sufficient parallelism is available the next challenge is preventing excess
parallelism and machine resource starvation, which is usually acheived with a thread pool or thread limit.

@@

\begin{description}
\item[Untracked dependencies] Some tasks depend on untracked values -- for example
C compilation will explicitly list the \cmd{source.c} file as a dependency,
but it may not record that the version of \cmd{gcc} is also a dependency.

\item[Non-determinism] Some tasks are \emph{non-deterministic}, producing any
result from a possible set. As an example, GHC programs compiled using parallelism
can change the order in which unique variables are obtained from the supply,
producing different but semantically identical results.

\item[Volatility] Some tasks are defined to change in every build, e.g. \Excel
provides a~``function'' \cmd{RANDBETWEEN} producing a fresh random number in a
specified range on each recalculation. Similarly, build systems like \Make and
\Shake provide volatile \emph{phony rules}. The main difference from
non-deterministic tasks is that volatile tasks cannot be cached. They are best
modelled as depending on a special key \cmd{RealWorld} whose value is changed
in every build.
\end{description}

@@

\todo{AM}{Show implementation of \hs{computeND}}.

@@

\begin{description}
\item[Communication] When traces or contents are stored in the cloud,
communication can become a bottleneck, so it is important to send only the
minimum amount of information, optimising with respect to build system
invariants. For example, incremental data processing systems in the cloud, such
as \Reflow~\cite{reflow}, need to efficiently orchestrate terabytes of data.

\item[Offloading] Once the cloud is storing build products and traces, it is
possible for the cloud to also contain dedicated workers that can execute tasks
remotely~--~offloading some of the computation and potentially running vastly
more commands in parallel.

\item[Eviction] The cloud storage, as modelled
in~\S\ref{sec-constructive-traces}, grows indefinitely, but often resource
constraints require evicting old items from the store. When evicting an old
value \hs{v}, one can also evict all traces mentioning the now-defunct
\hs{hash}~\hs{v}. However, for shallow builds (see below), it is beneficial to
keep these traces, allowing builds to ``pass-through'' hashes whose underlying
values are not known, recreating them only when they must be materialised.

\item[Shallow builds] Building the end result, e.g. an installer package, often
involves many intermediate tasks. The intermediate results may be large, so some
cloud build systems are designed to build end products \emph{without
materialising} intermediate results, producing a so-called \emph{shallow build}
(see an example in~\S\ref{sec-background-bazel}). Some build systems go even
further, integrating with the file system to only materialise the file when the
user accesses it \cite{gvfs}.
\end{description}

@@

\begin{figure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-build.pdf}}
% \vspace{-1.5mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-clean.pdf}}
% \vspace{-1.5mm}
\caption{Clean up, evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.34\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-rebuild.pdf}}
% \vspace{-1.5mm}
\caption{Build \cmd{main.prof} \& \cmd{report.txt}}
\end{subfigure}
% \vspace{-1.5mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool, then record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash value), then build \cmd{report.txt} by downloading it from the
matching deep constructive trace in the cloud, resulting in a Frankenbuild
because \cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted
cloud storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
% \vspace{-4mm}
\end{figure}

@@

\todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote execution service.}

@@

For examples like \LaTeX~we consider the proper encoding to not be circular
tasks, but a series of iterative steps, as described
by~\cite{shake-fixed-point}. It is important that the number of executions is
bounded, otherwise the build system may not terminate (a legitimate concern
with \LaTeX, which can be put into a situation where it is bistable or diverging
over multiple executions). The examples in \Excel tend to encode either mutable
state, or recurrence relations. The former is only required because \Excel
inherently lacks the ability to write mutable state, and the latter is probably
better solved using explicit recurrence formulae.

Overall we choose not to deal with cyclic dependencies -- a choice that most build
systems also follow. There are computation frameworks that support tasks with
cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
certain sense, e.g. see~\cite{pottier2009lazy} and~\cite{radul2009propagation}.

@@

Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
system that works directly on \hs{k} and \hs{v} values. However, some build
systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
values, allowing types that are stored only in the \Shake's build information.
\Shake users have remarked that polymorphism provides a much easier expression
of concepts, e.g. see~\cite{hadrian}.

As one example of richer key/value types, consider the version of
\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
to define an \emph{oracle} rule as per~\cite{mitchell2012shake} whose value is
the result of running \cmd{gcc -}\cmd{-version} and which is volatile, making the
\cmd{gcc} version something that can be depended upon. Of course, provided the
build can express volatile dependencies and supports cutoff, the version number
could equally be written to a file and used in a similar way, albeit at the cost
of introducing additional correctness invariants, e.g. that the file
\cmd{gcc-version.txt} does indeed contains two dot-separated integers.

Generalised algebraic data types (GADTs)~\cite{spj2006gadts} can be used to
extend our models to support ``typed'' build tasks. The idea is to replace
the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
more polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The idea is best explained by way of an example:

@@

\todo{AM/NM}{We are talking about sets above, but are using lists, which is a
bit inconsistent. We could switch from \hs{[k]} to \hs{Set k} fairly easily.
Shall we?}

>>>>

\item The \hs{restarting} scheduler can be made parallel in a few ways, but the
most direct is to have $n$ threads reading keys from the build queue. As before,
if a key requires a dependency not yet built, it is moved to the end~--~the
difference is that sometimes keys will be moved to the back of the queue not
because they are out of date but because of races with earlier tasks that had
not yet finished. As a consequence, if a the calculation order is persisted over
successive runs (as \Excel does), potentially racey dependencies will be
separated, giving better parallelism over time.

\item The \hs{suspending} scheduler can be made parallel by starting multiple
dependencies in parallel. One approach is to make the request for dependencies
take a list of keys, as implemented by \Shake. Another approach is to treat the
\hs{Applicative} dependencies of a \hs{Task}~\hs{Monad} in parallel, as
described by~Marlow~\etal~\shortcite{marlow2014haxl}.
\end{enumerate}

Once sufficient parallelism is available the next challenge is preventing excess
parallelism and machine resource starvation, which is usually achieved with a
thread pool or thread limit.

@@

\begin{itemize}
\item \textbf{Untracked dependencies}: Some tasks depend on untracked values --
      for example C compilation will explicitly list the \cmd{source.c} file as
      a dependency, but it may not record that the version of \cmd{gcc} is also
      a dependency.

\item \textbf{Non-determinism}: Some tasks are \emph{non-deterministic},
      producing any result from a possible set. As an example, GHC programs
      compiled using parallelism can change the order in which unique variables
      are obtained from the supply, producing different but semantically
      identical results.

\item \textbf{Volatility}: Some tasks are defined to change in every build, e.g.
      \Excel provides a~``function'' \cmd{RANDBETWEEN} producing a fresh random
      number in a specified range on each recalculation. Similarly, build
      systems like \Make and \Shake provide volatile \emph{phony rules}. The
      main difference from non-deterministic tasks is that volatile tasks cannot
      be cached. They are best modelled as depending on a special key
      \cmd{RealWorld} whose value is changed in every build.
\end{itemize}

@@

% \todo{AM}{Show implementation of \hs{computeND}.
% Neil says: not convinced it's that interesting.}.

@@

\begin{itemize}
\item \textbf{Communication}: When traces or contents are stored in the cloud,
communication can become a bottleneck, so it is important to send only the
minimum amount of information, optimising with respect to build system
invariants. For example, incremental data processing systems in the cloud, such
as \Reflow~\cite{reflow}, need to efficiently orchestrate terabytes of data.

\item \textbf{Offloading}: Once the cloud is storing build products and traces,
it is possible for the cloud to also contain dedicated workers that can execute
tasks remotely~--~offloading some of the computation and potentially running
vastly more commands in parallel.

\item \textbf{Eviction}: The cloud storage, as modelled
in~\S\ref{sec-constructive-traces}, grows indefinitely, but often resource
constraints require evicting old items from the store. When evicting an old
value \hs{v}, one can also evict all traces mentioning the now-defunct
\hs{hash}~\hs{v}. However, for shallow builds (see below), it is beneficial to
keep these traces, allowing builds to ``pass-through'' hashes whose underlying
values are not known, recreating them only when they must be materialised.

\item \textbf{Shallow builds}: Building the end result, e.g. an installer
package, often involves many intermediate tasks. The intermediate results may be
large, so some cloud build systems are designed to build end products
\emph{without materialising} intermediate results, producing a so-called
\emph{shallow build} (see an example in~\S\ref{sec-background-bazel}). Some
build systems go even further, integrating with the file system to only
materialise the file when the user accesses it~\cite{gvfs}.
\end{itemize}

@@

\begin{figure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-build.pdf}}
% \vspace{-1.5mm}
\caption{Initial build}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-clean.pdf}}
% \vspace{-1.5mm}
\caption{Clean up, evict \cmd{main.prof}}
\end{subfigure}
\begin{subfigure}[b]{0.36\linewidth}
\centerline{\includegraphics[scale=0.26]{fig/frankenbuild-example-rebuild.pdf}}
% \vspace{-1.5mm}
\caption{Build \cmd{main.prof} and \cmd{report.txt}}
\end{subfigure}
% \vspace{-1.5mm}
\caption{A Frankenbuild example: (a)~build a human-readable profiling report for
\cmd{main.exe} from information dump \cmd{main.prof} produced by a profiling
tool, then record deep constructive traces in the cloud, (b)~remove built files
locally and evict \cmd{main.prof} from the cloud storage, (c)~build
\cmd{main.prof} by executing the profiler (profiling is non-deterministic, hence
the new hash value), then build \cmd{report.txt} by downloading it from the
matching deep constructive trace in the cloud, resulting in a Frankenbuild
because \cmd{main.prof} and \cmd{report.txt} are inconsistent. New and evicted
cloud storage entries are highlighted; file hashes are shown in circles.
\label{fig-frankenbuild}}
% \vspace{-4mm}
\end{figure}

@@

% \todo{AM/NM}{Re-using existing infrastructure: (a) key-value store, (b) remote
% execution service. Neil says: I think this is a stretch. It's a lot more nuanced
% as we saw from Bazel discussions so I'd probably skip it.}

@@

For examples like \LaTeX~we consider the proper encoding to not be circular
tasks, but a series of iterative steps, as described by
Mitchell~\shortcite{shake-fixed-point}. It is important that the number of
executions is bounded, otherwise the build system may not terminate (a
legitimate concern with \LaTeX, which can be put into a situation where it is
bistable or diverging over multiple executions). The examples in \Excel tend to
encode either mutable state, or recurrence relations. The former is only
required because \Excel inherently lacks the ability to write mutable state, and
the latter is probably better solved using explicit recurrence formulae.

Overall we choose not to deal with cyclic dependencies -- a choice that most build
systems also follow. There are computation frameworks that support tasks with
cyclic dependencies under the assumption that tasks are \emph{monotonic} in a
certain sense~\cite{pottier2009lazy,radul2009propagation}.

@@

Our build system abstraction assumes a \hs{k}/\hs{v} store, along with a build
system that works directly on \hs{k} and \hs{v} values. However, some build
systems provide greater flexibility, e.g. \Shake permits polymorphic keys and
values, allowing types that are stored only in the \Shake's build information.
\Shake users have remarked that polymorphism provides a much easier expression
of concepts, e.g. see~Mokhov~\etal~\shortcite{hadrian}.

As one example of richer key/value types, consider the version of
\cmd{gcc}~--~for many builds it should be a dependency. In \Shake it is possible
to define an \emph{oracle} rule as per Mitchell~\shortcite{mitchell2012shake}
whose value is the result of running \cmd{gcc -}\cmd{-version} and which is
volatile, making the \cmd{gcc} version something that can be depended upon. Of
course, provided the build can express volatile dependencies and supports
cutoff, the version number could equally be written to a file and used in a
similar way, albeit at the cost of introducing additional correctness
invariants, e.g. that the file \cmd{gcc-version.txt} does indeed contains two
dot-separated integers.

Generalised algebraic data types, or GADTs~\cite{spj2006gadts}, can be used to
extend our models to support ``typed'' build tasks. The idea is to replace
the callback \hs{fetch}~\hs{::}~\hs{k}~\hs{->}~\hs{f}~\hs{v} by its
more polymorphic equivalent \hs{fetch}~\hs{::}~\hs{k}~\hs{v}~\hs{->}~\hs{f}~\hs{v},
where~\hs{k}~\hs{v} is a GADT representing keys tagged by the type~\hs{v} of
corresponding values. The idea is best explained by way of an example:

@@

% \todo{AM/NM}{We are talking about sets above, but are using lists, which is a
% bit inconsistent. We could switch from \hs{[k]} to \hs{Set k} fairly easily.
% Shall we? Neil says: Haskell code dealing with sets is quite a lot more complex.
% People are used to being quite fuzzy around the difference - I'd leave it as is.
% Andrey says: Also, using \hs{Set v} doesn't seem possible, because we won't be
% able to relate \hs{v}s with corresponding \hs{k}s}.
